---
title: "6289_final_3"
author: "rballard myork [uname]"
date: "2025-04-15"
output: html_document
---


# PROJECT NAME

## Frontmatter
```{r ezidsetup, eval= FALSE}
#https://github.com/physicsland/ezids
if (!requireNamespace("remotes", quietly = TRUE)) { install.packages("remotes") }
install.packages ("devtools")
devtools::install_github("physicsland/ezids")
```
The above installs ezids, a package manager developed and maintained by GW DS Department.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(20052)

remotes::install_github("mikejohnson51/NFHL")
remotes::install_github("mikejohnson51/AOI")



library(ezids)
pkg_vector <- c("sf", "dplyr", "tmap", "terra", "NFHL", "exactextractr","spatialreg", "broom", "ggplot2","lmtest", "sf", "spData","spdep", "sphet", "INLA","NFHL","AOI")

invisible(lapply(pkg_vector,loadPkg))


```




# 1. FEMA Flood Zone Extraction

- Extract Layer 28 polygons for the DMV region using the `NFHL` R package.
- Filter for **Special Flood Hazard Areas (SFHA)** using the condition `SFHA_TF == "T"`.

# 2. NOAA Flood Event Collection

- Download and filter NOAA flood events (1950–2023) by event type:
  - Flood
  - Flash Flood
  - Coastal Flood
- Extract:
  - Geolocation
  - Event date
  - Event type
  - Damage estimates

# 3. Spatial Join and Aggregation

- Assign events to spatial units:
  - Census tracts or
  - ZIP Code Tabulation Areas (ZCTAs)
- Count flood event frequency per spatial unit.
- Overlay flood event hotspots on FEMA SFHA zones.
- Identify:
  - **Matching areas**: events occurring inside SFHA zones
  - **Mismatch areas**: events occurring outside SFHA zones

# 4. Time Series Analysis

- Aggregate flood events by decade.
- Plot trends in:
  - Event frequency
  - Severity using damage amounts