---
title: "6289_final_3"
author: "rballard"
date: "2025-04-15"
output: html_document
---


# PROJECT NAME

## Frontmatter
```{r ezidsetup,eval=FALSE}
#https://github.com/physicsland/ezids
if (!requireNamespace("remotes", quietly = TRUE)) { install.packages("remotes") }
install.packages ("devtools")
devtools::install_github("physicsland/ezids")
install.packages("ezids")
```
The above installs ezids, a package manager developed and maintained by GW DS Department.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

set.seed(20052)

#install.packages("rfema", repos = "https://ropensci.r-universe.dev")

library(ezids)
pkg_vector <- c("sf", "dplyr", "tmap", "terra", "NFHL", "exactextractr","spatialreg", "broom", "ggplot2","lmtest", "sf", "spData","spdep", "sphet", "INLA","NFHL","AOI", "tidycensus", "tigris","rfema","raster","stringr","xml2","rvest","httr","reader","purrr","readr","rvest","stringr","tidyr")

invisible(lapply(pkg_vector,loadPkg))

data_dir <- "C:/Users/russe/Documents/github/6289spatial_final/data/"
```


# 1. FEMA Flood Zone Extraction

- Extract Layer 28 polygons for the DMV region using the `NFHL` R package.
- Filter for **Special Flood Hazard Areas (SFHA)** using the condition `SFHA_TF == "T"`.

```{r createnfhldf}

# FEMA NFHL base URL
base_url <- "https://hazards.fema.gov/femaportal/NFHL/"
target_url <- paste0(base_url, "searchResult")

# Read page
page <- read_html(target_url)

# Extract all table rows
rows <- page %>% html_nodes("tr")

# Parse each row into fields: FIRMID, County, State, Date, Size, Link
nfhl_df <- rows %>%
  map_df(~{
    tds <- .x %>% html_nodes("td")
    if (length(tds) < 6) return(NULL)  # skip malformed rows

    list(
      dfirm_id = tds[1] %>% html_text(trim = TRUE),
      county = tds[2] %>% html_text(trim = TRUE),
      state = tds[3] %>% html_text(trim = TRUE),
      effective_date = tds[4] %>% html_text(trim = TRUE),
      file_size = tds[5] %>% html_text(trim = TRUE),
      url = tds[6] %>% html_node("a") %>% html_attr("href") %>% paste0(base_url, .)
    )
  }) %>%
  filter(str_detect(dfirm_id, "^(11|24|51).*")) %>%
  mutate(url = URLencode(url, reserved = TRUE))

# Preview
print(head(nfhl_df))

```

```{r nfhl_scrape, eval=FALSE}

# Base URL and scrape target
base_url <- "https://hazards.fema.gov/femaportal/NFHL/"
target_url <- paste0(base_url, "searchResult")

# Read and parse HTML
page <- read_html(target_url)

# Extract all download links with their file names
zip_links <- page %>%
  html_nodes("a") %>%
  html_attr("href") %>%
  na.omit() %>%
  str_subset("^Download/ProductsDownLoadServlet\\?DFIRMID=") %>%
  unique()

# Filter by DFIRMID prefix: 11, 24, or 51
filtered_links <- zip_links %>%
  str_subset("^Download/ProductsDownLoadServlet\\?DFIRMID=(11|24|51)")

# Convert to full URLs
full_urls <- paste0(base_url, filtered_links)
full_urls <- gsub(" ", "%20", full_urls)


# Optional: Set destination folder
dest_dir <- "data/test/nfhl_zips"
dir.create(dest_dir, showWarnings = FALSE)

# Download function with polite wait
download_nfhl_zip <- function(url, delay = 3) {
  file_name <- str_extract(url, "fileName=[^&]+") %>%
    str_remove("fileName=")
  dest_file <- file.path(dest_dir, file_name)

  if (!file.exists(dest_file)) {
    message("Downloading: ", file_name)
    tryCatch({
      download.file(url, destfile = dest_file, mode = "wb")
      Sys.sleep(delay)
    }, error = function(e) {
      message("Error downloading ", file_name)
    })
  } else {
    message("Already downloaded: ", file_name)
  }
}

# Loop and download
walk(full_urls, download_nfhl_zip)

```

The above codeblock scrapes FEMA's NFHL DFIRM6 catalog.

```{r create_geoidpivot}

library(sf)
library(dplyr)
library(stringr)
library(tidyr)
library(purrr)

# Directory containing your zip files
zip_dir <- "C:/Users/russe/Documents/github/6289spatial_final/data/test/nfhl_zips"

# List of zip files (assuming they start with 5-digit GEOID prefix)
zip_files <- list.files(zip_dir, pattern = "^\\d{5}.*\\.zip$", full.names = TRUE)

# Initialize output data frame
outputs <- tibble(GEOID = character(), FLD_ZONE = character(), COUNT = integer())

# Loop through each zip file
for (zip_path in zip_files) {
  # Extract the 5-digit GEOID from the filename
  geoid <- str_extract(basename(zip_path), "^\\d{5}")

  # Create temporary directory for extraction
  temp_dir <- tempfile(pattern = "nfhl_")
  dir.create(temp_dir)

  # Extract contents
  unzip(zip_path, exdir = temp_dir)

  # Look for the S_FLD_HAZ_AR.shp shapefile
  shp_files <- list.files(temp_dir, pattern = "S_FLD_HAZ_AR\\.shp$", recursive = TRUE, full.names = TRUE)
  if (length(shp_files) == 0) {
    message(glue::glue("No shapefile found in {zip_path}, skipping."))
    unlink(temp_dir, recursive = TRUE)
    next
  }

  # Load shapefile
  flood_layer <- tryCatch({
    st_read(shp_files[1], quiet = TRUE)
  }, error = function(e) {
    message(glue::glue("Failed to read shapefile in {zip_path}, skipping."))
    unlink(temp_dir, recursive = TRUE)
    return(NULL)
  })

  if (is.null(flood_layer)) next

  Summarize FLD_ZONE counts
  pivot <- flood_layer %>%
    st_drop_geometry() %>%
    count(FLD_ZONE, name = "COUNT") %>%
    mutate(GEOID = geoid) %>%
    dplyr::select(GEOID, FLD_ZONE, COUNT)
  

  # Append to outputs
  outputs <- bind_rows(outputs, pivot)

  
  # Cleanup
  unlink(temp_dir, recursive = TRUE)
}

write.csv(outputs, file = "data/outputs.csv", row.names = FALSE)

```

```{r load_outputs}
outputs <- read_csv("data/outputs.csv", col_types = cols(.default = "c"))
```

# 2. NOAA Flood Event Collection

- Download and filter NOAA flood events (1950–2023) by event type:
  - Flood
  - Flash Flood
  - Coastal Flood

```{r noaa_bulkdatadownload, eval=FALSE}
#This code chunk will download all the NOAA event data so don't run if data is already present in environment.

# Base URL of the directory
base_url <- "https://www.ncei.noaa.gov/pub/data/swdi/stormevents/csvfiles/"

# Read the HTML content of the directory listing
page <- read_html(base_url)

# Extract all links to .csv.gz files
file_links <- page %>%
  html_nodes("a") %>%
  html_attr("href") %>%
  na.omit() %>%
  str_subset("\\.csv\\.gz$")

# Make full URLs
full_urls <- paste0(base_url, file_links)

# Download folder (you can change this)
download_dir <- "C:/Users/russe/Documents/github/6289spatial_final/data/test"
dir.create(download_dir, showWarnings = FALSE)

# Loop to download with delays
for (url in full_urls) {
  file_name <- basename(url)
  dest_path <- file.path(download_dir, file_name)
  
  message("Downloading: ", file_name)
  tryCatch({
    download.file(url, destfile = dest_path, mode = "wb")
  }, error = function(e) {
    warning("Failed to download ", file_name, ": ", conditionMessage(e))
  })
  
  # Wait a few seconds to avoid hammering the server
  Sys.sleep(runif(1, 2, 5))  # random wait between 2 and 5 seconds
}

```

The above code chunk downloads NOAA Event Details for 1950-2024. It should only be run if data is not already present in development environment.

```{r noaa_rowjoin}
#rowwise joins of event details, fatalities, and locations


files <- list.files(data_dir, pattern = "\\.csv\\.gz$", full.names = TRUE)

details_files <- files[grepl("StormEvents_details", files)]
locations_files <- files[grepl("StormEvents_locations", files)]
fatalities_files <- files[grepl("StormEvents_fatalities", files)]

read_as_character <- function(file) {
  df <- read_csv(file, col_types = cols(.default = col_character()), show_col_types = FALSE)
  df$filename <- basename(file)
  df
}

# Combine all details files
storm_details_df <- map_dfr(details_files, read_as_character)

# Combine all locations files
storm_locations_df <- map_dfr(locations_files, read_as_character)

# Combine all fatalities files
storm_fatalities_df <- map_dfr(fatalities_files, read_as_character)

head(storm_details_df)

```

The above code chunk reads in and conducts a rowwise join of the annual NOAA event tables 

```{r noaa_tablewrite, eval=FALSE}
#writes processed table

write_pipe_csv_gz <- function(df, output_dir, filename) {
  # Ensure the output directory exists
  dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
  
  # Construct full path
  file_path <- file.path(output_dir, paste0(filename, ".csv.gz"))
  
  # Write the compressed, pipe-delimited file
  readr::write_delim(
    df,
    file = file_path,
    delim = "|",
    col_names = TRUE
  )
  
  message("File written to: ", file_path)
}

write_pipe_csv_gz(storm_details_dmv, data_dir, "storm_details_dmv")
```

The above chunk will write the joined noaa event details table to data_dir.


```{r noaa_tableload}
#will read the filtered storm_details_dmv.csv.gz saved to your repo data directory from GDrive

data_dir = "C:/Users/russe/Documents/github/6289spatial_final/data"

read_pipe_csv_gz <- function(input_dir, filename) {
  file_path <- file.path(input_dir, paste0(filename, ".csv.gz"))
  
  df <- readr::read_delim(
    file = file_path,
    delim = "|",
    col_names = TRUE,
    show_col_types = FALSE,
  col_types = cols(.default = col_character())
  )
  
  return(df)
}

storm_details_df <- as_tibble(read_pipe_csv_gz(data_dir, "storm_details_dmv"))
```

The above code chunk can be used to load a previously saved noaa event details .csv.gz to proceed with processing.

```{r noaa_detailsfilter}
#filters event details to include only DC, MD, VA state fips codes
storm_details_df <- storm_details_df %>%
  filter(STATE_FIPS %in% c("11", #DC
                           "24", #MD
                           "51"  #VA
                           ))

storm_details_df <- storm_details_df %>%
  filter(EVENT_TYPE %in% c('Flash Flood',
                           'Flood',
                           'Coastal Flood'
                           ))
```

The above code chunk filters the events data to include only STATE_FIPS for DC/MD/VA and to only include EVENT_TYPES of Coastal Flood, Flash Flood, or Flood.


- Extract:
  - Geolocation
  - Event date
  - Event type
  - Damage estimates

```{r noaatablesubset}

storm_details_df <- storm_details_df %>%
  dplyr::select("EVENT_ID",
                "BEGIN_YEARMONTH",
                "BEGIN_DAY",
                "BEGIN_TIME",
                "STATE",
                "STATE_FIPS",
                "YEAR",
                "MONTH_NAME",
                "EVENT_TYPE",
                "CZ_FIPS",
                "CZ_NAME",
                "BEGIN_DATE_TIME",
                "INJURIES_DIRECT",
                "INJURIES_INDIRECT",
                "DEATHS_DIRECT",
                "DEATHS_INDIRECT",
                "DAMAGE_PROPERTY",
                "DAMAGE_CROPS",
                "FLOOD_CAUSE",
                "BEGIN_LOCATION",
                "BEGIN_LAT",
                "BEGIN_LON"
)


```

```{r noaaevents_cleaning}
storm_details_df <- storm_details_df %>%
  mutate(
    DAMAGE_CROPS_CLEAN = case_when(
      str_detect(DAMAGE_CROPS, regex("^[0-9.]+K$", ignore_case = TRUE)) ~
        as.numeric(str_remove(DAMAGE_CROPS, regex("K", ignore_case = TRUE))) * 1e3,
      str_detect(DAMAGE_CROPS, regex("^[0-9.]+M$", ignore_case = TRUE)) ~
        as.numeric(str_remove(DAMAGE_CROPS, regex("M", ignore_case = TRUE))) * 1e6,
      str_detect(DAMAGE_CROPS, "^[0-9.]+$") ~
        as.numeric(DAMAGE_CROPS),
      TRUE ~ NA_real_
    )
  )

storm_details_df <- storm_details_df %>%
  mutate(
    DAMAGE_PROPERTY_CLEAN = case_when(
      str_detect(DAMAGE_PROPERTY, regex("^[0-9.]+K$", ignore_case = TRUE)) ~
        as.numeric(str_remove(DAMAGE_PROPERTY, regex("K", ignore_case = TRUE))) * 1e3,
      str_detect(DAMAGE_PROPERTY, regex("^[0-9.]+M$", ignore_case = TRUE)) ~
        as.numeric(str_remove(DAMAGE_PROPERTY, regex("M", ignore_case = TRUE))) * 1e6,
      str_detect(DAMAGE_PROPERTY, "^[0-9.]+$") ~
        as.numeric(DAMAGE_PROPERTY),
      TRUE ~ NA_real_
    )
  )


storm_details_df <- storm_details_df %>%
  mutate(across(c(INJURIES_DIRECT, INJURIES_INDIRECT, DEATHS_DIRECT, DEATHS_INDIRECT), as.numeric))


storm_details_df <- storm_details_df %>%
  mutate(
    FIPS5 = paste0(STATE_FIPS, str_pad(CZ_FIPS, width = 3, pad = "0"))
  )

storm_details_df <- storm_details_df %>%
  mutate(
    TOTAL_DAMAGE = DAMAGE_PROPERTY_CLEAN + DAMAGE_CROPS_CLEAN
  )
```


# 3. Spatial Join and Aggregation

- Assign events to spatial units:
  - Census tracts or
  - ZIP Code Tabulation Areas (ZCTAs)
  
```{r constructlinktable}
library(tigris)
library(dplyr)

# Get all U.S. counties (this pulls GEOID, NAME, STATEFP, COUNTYFP)
county_sf <- counties(cb = TRUE, year = 2023)

# Just keep needed fields and convert to character
fips_df <- county_sf %>%
  st_drop_geometry() %>%
  transmute(
    NAME,
    STATEFP,
    COUNTYFP,
    FIPS5 = paste0(STATEFP, COUNTYFP),
    GEOID = GEOID  # This should match FIPS5 normally
  )

# Your FEMA output GEOIDs (e.g., from the earlier script)
fema_geoids <- outputs %>%
  distinct(GEOID) %>%
  pull()

# Identify GEOIDs that are in FEMA data but not in official FIPS5
fema_geoids_df <- tibble(FEMA_GEOID = fema_geoids)

lookup_table <- fema_geoids_df %>%
  anti_join(fips_df, by = c("FEMA_GEOID" = "FIPS5"))

manual_fixes <- tibble(
  FEMA_DFIRM = c("11000", #DC City
                 "24000", #Baltimore City
                 "51002", #Bristol City
                 "51004", #Cumberland County, there are two shapefile sets for this data, also exists as 51049, fips 51049
                 "51006", #Franklin City
                 "51008", #City of Hopewell
                 "51010", #City of Norfolk
                 "51012", #City of Radford
                 "51018", # City of Poquoson
                 "51551", # City of Alexandria),
                 "51540"), #City of Charlottesville
  FIPS5 = c("11001", #DC City
            "24005", #Baltimore City
            "51502", #Bristol City
            "51049", #Cumberland County
            "51620", #Franklin City
            "51670", #City of Hopewell
            "51710 ", #City of Norfolk
            "51750", #City of Radford
            "51735", #City of Poquoson
            "51510", #City of Alexandria
            "51003"
            ))

```

```{r manualfix_nhfl}

outputs2 <- outputs %>%
  left_join(manual_fixes, by = c("GEOID" = "FEMA_DFIRM")) %>%
  mutate(GEOID = if_else(!is.na(FIPS5), FIPS5, GEOID)) %>%
  dplyr::select(-FIPS5)

outputs2$COUNT <- as.numeric(outputs$COUNT)

outputs_wide <- outputs2 %>%
  group_by(GEOID, FLD_ZONE) %>%
  summarise(COUNT = sum(COUNT, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = FLD_ZONE,
    values_from = COUNT,
    values_fill = list(COUNT = 0)
  )

```

```{r join_nfhlouttonoaaevent}

library(dplyr)

events_flood_df <- storm_details_df %>%
  left_join(outputs_wide, by = c("FIPS5" = "GEOID"))

subset_na_A <- events_flood_df %>%
  filter(is.na(A)) %>%
  dplyr::select(FIPS5, STATE, CZ_NAME)

distinct_na_A_FIPS5 <- events_flood_df %>%
  filter(is.na(A)) %>%
  dplyr::select(FIPS5, STATE, CZ_NAME) %>%
  distinct(FIPS5, .keep_all = TRUE)

```

```{r write_eventsflood}
write.csv(events_flood_df, file = "data/events_flood_df.csv", row.names = FALSE)
```

```{r load_eventsflood}
events_flood_df <- read_csv("data/events_flood_df.csv", col_types = cols(.default = "c"))

events_flood_df <- events_flood_df %>%
  mutate(
    YEAR = as.numeric(as.character(YEAR)),
    DAMAGE_PROPERTY_CLEAN = as.numeric(as.character(DAMAGE_PROPERTY_CLEAN)),
    DAMAGE_CROPS_CLEAN = as.numeric(as.character(DAMAGE_CROPS_CLEAN)),
    DEATHS_DIRECT = as.numeric(as.character(DEATHS_DIRECT)),
    DEATHS_INDIRECT = as.numeric(as.character(DEATHS_INDIRECT)),
    INJURIES_DIRECT = as.numeric(as.character(INJURIES_DIRECT)),
    INJURIES_INDIRECT = as.numeric(as.character(INJURIES_INDIRECT))
  )

```

- Count flood event frequency per spatial unit.
- Overlay flood event hotspots on FEMA SFHA zones.
- Identify:
  - **Matching areas**: events occurring inside SFHA zones
  - **Mismatch areas**: events occurring outside SFHA zones

# 4. Time Series Analysis

- Aggregate flood events by decade.

```{r transform_eventsflood_decadeadd}
events_flood_df <- events_flood_df %>%
  mutate(
    DECADE = case_when(
      YEAR >= 1990 & YEAR < 2000 ~ "90s",
      YEAR >= 2000 & YEAR < 2010 ~ "00s",
      YEAR >= 2010 & YEAR < 2020 ~ "10s",
      YEAR >= 2020 & YEAR < 2030 ~ "2020-2024",
      TRUE ~ "Other"
    )
  ) %>%
  mutate(
    DECADE = factor(DECADE, levels = c("90s", "00s", "10s", "2020-2024"))
  )
```

- Plot trends in:
  - Event frequency
  - Severity using damage amounts

```{r pivot_eventsflood_fips5decade}

decade_frequency_f <- events_flood_df %>%
  group_by(FIPS5, DECADE) %>%
  summarise(
    UNIQUE_EVENTS = n_distinct(EVENT_ID),
    .groups = "drop"
  )

# Summarize total damages by year and state
decade_damage_f <- events_flood_df %>%
  group_by(FIPS5,DECADE) %>%
  summarise(
    TOTAL_DAMAGE = sum(DAMAGE_PROPERTY_CLEAN + DAMAGE_CROPS_CLEAN, na.rm = TRUE),
    .groups = "drop")

decade_casualty_f <- events_flood_df %>%
  group_by(FIPS5,DECADE) %>%
  summarise(
    TOTAL_CASUALTIES = sum(
      DEATHS_DIRECT + DEATHS_INDIRECT + INJURIES_DIRECT + INJURIES_INDIRECT,
      na.rm = TRUE),
    .groups = "drop"
  )

```

```{r pivot_eventsflood_totaldecade}
decade_frequency_t <- events_flood_df %>%
  group_by(DECADE) %>%
  summarise(
    UNIQUE_EVENTS = n_distinct(EVENT_ID),
    .groups = "drop"
  )

decade_damage_t <- events_flood_df %>%
  group_by(DECADE) %>%
  summarise(
    UNIQUE_EVENTS = n_distinct(EVENT_ID),
    .groups = "drop")

decade_casualty_t <- events_flood_df %>%
  group_by(DECADE) %>%
  summarise(
    TOTAL_CASUALTIES = sum(
      DEATHS_DIRECT + DEATHS_INDIRECT + INJURIES_DIRECT + INJURIES_INDIRECT,
      na.rm = TRUE),
    .groups = "drop"
  )
```


```{r viz_eventsflood_decadedamage}
dmg_fips <- ggplot(decade_damage, aes(x = DECADE, y = log(TOTAL_DAMAGE+1))) +
  geom_boxplot(outlier.alpha = 0.3, fill = "#4682b4") +
  labs(
    title = "Distribution of Total Flood Event Damage\nby FIPS5 (County) by Decade",
    x = "Decade",
    y = "Log Total Flood Event Damage by FIPS5 (County)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size = 12)
  )

dmg_tot <- ggplot(decade_frequency_t, aes(x = DECADE, y = log(TOTAL_DAMAGE + 1))) +
  geom_col(fill = "#4682b4") +
  labs(
    title = "Total Flood Damage by Decade",
    x = "Decade",
    y = "Log Total Flood Damage"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size = 12)
  )

library(gridExtra)
grid.arrange(dmg_fips, dmg_tot, ncol = 2, widths = c(1, 1))
```

```{r viz_eventsflood_decadefrequency}

freq_fips <- ggplot(decade_frequency_f, aes(x = DECADE, y = UNIQUE_EVENTS)) +
  geom_boxplot(outlier.alpha = 0.3, fill = "#4682b4") +
  labs(
    title = "Distribution of Total Flood Events by\nFIPS5 (County) by Decade",
    x = "Decade",
    y = "Total Flood Events by FIPS5 (County)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size = 12)
  )

freq_tot <- ggplot(decade_frequency_t, aes(x = DECADE, y = UNIQUE_EVENTS)) +
  geom_col(fill = "#4682b4") +
  labs(
    title = "Distribution of Total Flood Events by\nFIPS5 (County) by Decade",
    x = "Decade",
    y = "Total Flood Events"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size = 12)
  )

grid.arrange(freq_fips, freq_tot, ncol = 2, widths = c(1, 1))
```

```{r viz_eventsflood_decadecasualty}

caus_fips <- ggplot(decade_casualty_f, aes(x = DECADE, y = TOTAL_CASUALTIES)) +
  geom_boxplot(outlier.alpha = 0.3, fill = "#4682b4") +
  labs(
    title = "Distribution of Total Flood Casualties by\nFIPS5 (County) by Decade",
    x = "Decade",
    y = "Total Flood Events by FIPS5 (County)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size = 12)
  )

caus_tot <- ggplot(decade_casualty_t, aes(x = DECADE, y = TOTAL_CASUALTIES)) +
  geom_col(fill = "#4682b4") +
  labs(
    title = "Distribution of Total Flood Casualties by Decade",
    x = "Decade",
    y = "Total Flood Casualties"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size = 12)
  )

grid.arrange(freq_fips, freq_tot, ncol = 2, widths = c(1, 1))
```
